Compositional Certified Resource Bounds\ifshort{\vspace{-1.15cm}}{}

### 2015-01-05 paper/main.tex
remove references to resource metrics

The derivation of a resource bound using potential functions is best
explained by example.  If we use the tick metric that assigns cost $n$
to the function call \code{tick(n)} and cost $0$ to all other
operations then the cost of the following example can be bounded by
$|[x,y]| = \max(y{-}x,0)$.% \footnote{Note that there are no restrictions on
  % the signs of the input variables in the examples.}
\begin{lstlisting}[basicstyle=\tt\small]
while (x<y) { x=x+1; tick(1); } #\hfill \textnormal{(Example 1)}#
\end{lstlisting}
To derive this bound, we start with the initial potential $\Phi_0 =
|[x,y]|$, which
we also use as the loop invariant.  For the loop body we have (like in
Hoare logic) to derive a triple like
$\htriple{\Phi_0}{\code{x=x+1;tick(1)}}{\Phi_0}$.  We can only do so
if we utilize the fact that $x<y$ at the beginning of the loop body.
The reasoning then works as follows. We start with the potential
$|[x,y]|$ and the fact that $|[x,y]| > 0$ before the assignment.  If
we denote the updated version of $x$ after the assignment by $x'$ then
the relation $|[x,y]| = |[x',y]| + 1$ between the potential before and
after the assignment \code{x=x+1} holds.  This means that we have the
potential $|[x,y]| + 1$ before the statement \code{tick(1)}.  Since
\code{tick(1)} consumes one resource unit, we end up with potential
$|[x,y]|$ after the loop body and have established the loop invariant
again.

@@

The concept of a potential function is a generalization of the
concept of a ranking function.  A potential function can be used like
a ranking function if we use the tick metric and add the statement
\code{tick(1)} to every back edge of the program (loops and function
calls).  However, a potential function is more flexible.  For example,
we can use a potential function to prove that Example~2 does not
consume any resources in the tick metric.
\begin{lstlisting}[basicstyle=\tt\small]
#\!\!\!\!\!#while (x<y) {tick(-1); x=x+1; tick(1)} #\hfill \textnormal{(Example 2)}#

@@

\begin{figure}[t]
\begin{minipage}[b]{0.8\linewidth}
\begin{lstlisting}[numbers=left]
${\color{blue}%
  \mbox{logical state invariant } \{ na = \#_1(a)} \}$
while (k > 0) {
  x=0;
  while (x < N && a[x] == 1) {
    @assert(na > 0);@
    a[x]=0; @na--;@
    tick(1); x++; }
  if (x < N) { a[x]=1; @na++;@ tick(1); }
  k--;
}
\end{lstlisting}
\vspace{-.2cm}
\end{minipage}
\caption{Assisted bound derivation using logical state.  We write
  $\#_1(a)$ for $\# \{ i \mid 0 {\le} i {<} N {\land} a[i] {=} 1 \}$ and use the
  tick metric.  The derived bound  is
  $2|[0,k]| + |[0,na]|$.
  }
\label{fig:xmplincaux}
\end{figure}

\begin{figure}
\vspace{-.3cm}
\begin{lstlisting}[numbers=left]
${\color{blue}%
  \mbox{logical state invariant }
  \{ lg > \log_2(h - l) \} }$
bsearch(x,l,h@,lg@) {
  if (h-l > 1) {
    @assert(lg > 0);@
    m = l + (h-l)/2;
    @lg--;@ if (a[m]>x) h=m; else l=m;
    tick($M_\code{bsearch}$);
    l = bsearch(x,l,h@,lg@);
    tick($-M_\code{bsearch}$);
  } else return l;
}
\end{lstlisting}
\vspace{-.2cm}
\caption{Assisted bound derivation using logical state.
  % We use the tick metric to model the stack consumption of the program.
  % Before a function call to \code{bsearch}, a tick statement pays for
  % the space used by stack frame of the function.  The space is recoverd
  % after the call.
  We write $\log_2(x)$ for the integer part of
  logarithm of $x$ in base 2.
  The semi-automatically derived bound is $|[0,lg]|$.
  }
\label{fig:xmplbsaux}
\end{figure}

\ifdefined\fullversion
\begin{figure}
\begin{lstlisting}
${\color{blue}
  \mbox{logical state invariant }
  \{ nm = A(0) \} }$
i=1; m=a[0];
while (i < N) {
  if (a[i] > m) {
    @assert(nm > 0); nm--;@
    m = a[i], tick(1);
  } i++, tick(1);
}
\end{lstlisting}
\caption{Example of assisted bound derivation using logical state.
  We write $A(i)$ for
  $\#\{ k \mid i \le k \le N \land \forall\, 0\le j<k.\, a[j] < a[k]\}$.
  We use the tick metric.  The bound derived is $|[0,N]| + |[0,nm]|$.
  }
\label{fig:xmplmax}
\end{figure}
\fi

@@

To compare our tool with existing work, we focused on loop bounds and
use a simple metric that counts the number of back edges (i.e., number
of loop iterations) that are followed in the execution of the program
because most other tools only bound this specific cost. In
\pref{fig:compar}, we show the bounds we derived ($\toolname$) together with
the bounds derived by LOOPUS~\cite{SinnZV14} and
Rank~\cite{AliasDFG10}.  We also contacted the authors of SPEED but
have not been able to obtain this tool.
KoAT~\cite{BrockschmidtEFFG14} and PUBS~\cite{AlbertAGPZ12} currently
cannot operate on C code and the examples would need to be manually
translated into a term-rewriting system to be analyzed by these tools.  For
Rank it is not completely clear how the computed bound relates to the C
program since the computed bound is for transitions in an automaton
that is derived from the C code.  For instance, the bound $2+y-x$ that
is derived for \emph{t08} only applies to the first loop in the
program.

@@

\pref{tab:cbench} contains a compilation of the results of our
experiments with the cBench benchmark suite. It shows a representative
list of automatically derived function bounds. In total we analyzed
more than $\totalLoc$ lines of code.  In the LoC column
we not only count the lines of the analyzed function but
also the ones of all the function it calls. \iffull{The line count is
computed on pre-processed C with all unnecessary declarations removed.}{}
We analyzed the functions using a metric that assigns a cost 1 to all
the back-edges in the control flow (loops, and function calls).
The bounds for the functions \code{ycc\_rgb\_conv} and \code{uv\_decode}
have been inferred with user interaction as described in \pref{sec:anno}.
The most challenging functions for $\toolname$ have unrolled loops
where many variables are assigned.  This stresses our analysis because
the number of LP variables has a quadratic growth in program variables.
Even on these stressful examples, the analysis could finish in less
than 2 seconds.  \iffull{These times can be reduced by adding
manual annotations to indicate $\toolname$ that certain variables do
not have to be tracked.  We could also envision to implement heuristics.}{}
%
For example, the \code{sha\_update} function
is composed of one loop calling two helper functions that in turn
have 6 and 1 inner loops.  In the analysis of the SHA algorithm,
the compositionality of our analysis is essential to get a tight bound
since loops on the same index are sequenced 4 and 2 times without
resetting it.  All other tools derive much larger constant factors.

@@

In this appendix we provide a non-exhaustive catalog of classes of
programs that can be automatically analyzed by our system.  For
simplicity, and to compare our analysis with existing tools, we use a
cost metric that counts the number of loop iterations and function
calls.  This is the only cost metric supported by tools like
KoAT~\cite{BrockschmidtEFFG14} and LOOPUS~\cite{SinnZV14}.  Sometimes
we also use the \emph{tick metric} if we want to discuss features such as
resource restitution.  The tick metric assigns cost $n$ to the
statement \code{tick(n)} and cost $0$ to all other statements. Of
course, the examples can also be analyzed with any other cost metric
in our system.

@@

Example \emph{ex1} from~\cite{GulwaniZ10} specifically focuses on the
code in the \emph{if} statement.  So we use the \emph{tick metric} and
insert \code{tick(1)} inside the if statement to derive a bound on
the number of times the code in the if statement is executed.  Note
that we cannot derive a bound for the whole program since the outer
loop is executed a quadratic number of times.  Nevertheless it is
straightforward to derive a bound on the number of ticks using the
amortized approach: In the if statement we know that $n>0$ and assign
\code{n=n-1}.  So we can use the potential of the interval size $|[0,n]|$
to pay for the tick.

@@

Example \emph{t46} shows a program that uses and returns resources.
Again, we use the \emph{tick metric} and the function \code{tick} to
describe the resource usage.  The function \code{produce} produces
$|[0,x]|$ resources, that is, in each of the $|[0,x]|$ iterations, it
receives one resource unit.  Similarly, the function \code{consume}
consumes $|[0,y]|$ resources.  The analyzer computes the tight bound
$|[0,y]|$ for the function \code{main}.  This is only possible since
amortized analysis naturally tracks the size changes to the variables
$x$ and $y$, and the interaction between \code{consume} and
\code{produce}.

>>>>

The derivation of a resource bound using potential functions is best
explained by example.  If we use a cost model that assigns cost $n$ to
the function call \code{tick(n)} and cost $0$ to all other operations
then the cost of the following example can be bounded by
$|[x,y]| = \max(y{-}x,0)$.% \footnote{Note that there are no restrictions on
  % the signs of the input variables in the examples.}
\begin{lstlisting}[basicstyle=\tt\small]
while (x<y) { x=x+1; tick(1); } #\hfill \textnormal{(Example 1)}#
\end{lstlisting}
To derive this bound, we start with the initial potential $\Phi_0 =
|[x,y]|$, which
we also use as the loop invariant.  For the loop body we have (like in
Hoare logic) to derive a triple like
$\htriple{\Phi_0}{\code{x=x+1;tick(1)}}{\Phi_0}$.  We can only do so
if we utilize the fact that $x<y$ at the beginning of the loop body.
The reasoning then works as follows. We start with the potential
$|[x,y]|$ and the fact that $|[x,y]| > 0$ before the assignment.  If
we denote the updated version of $x$ after the assignment by $x'$ then
the relation $|[x,y]| = |[x',y]| + 1$ between the potential before and
after the assignment \code{x=x+1} holds.  This means that we have the
potential $|[x,y]| + 1$ before the statement \code{tick(1)}.  Since
\code{tick(1)} consumes one resource unit, we end up with potential
$|[x,y]|$ after the loop body and have established the loop invariant
again.

@@

The concept of a potential function is a generalization of the
concept of a ranking function.  A potential function can be used like
a ranking function if we add the statement
\code{tick(1)} to every back edge of the program (loops and function
calls).  However, a potential function is more flexible.  For example,
we can use a potential function to prove that Example~2 does not
consume any resources.
\begin{lstlisting}[basicstyle=\tt\small]
#\!\!\!\!\!#while (x<y) {tick(-1); x=x+1; tick(1)} #\hfill \textnormal{(Example 2)}#

@@

\begin{figure}[t]
\begin{minipage}[b]{0.8\linewidth}
\begin{lstlisting}[numbers=left]
${\color{blue}%
  \mbox{logical state invariant } \{ na = \#_1(a)} \}$
while (k > 0) {
  x=0;
  while (x < N && a[x] == 1) {
    @assert(na > 0);@
    a[x]=0; @na--;@
    tick(1); x++; }
  if (x < N) { a[x]=1; @na++;@ tick(1); }
  k--;
}
\end{lstlisting}
\vspace{-.2cm}
\end{minipage}
\caption{Assisted bound derivation using logical state.  We write
  $\#_1(a)$ for $\# \{ i \mid 0 {\le} i {<} N {\land} a[i] {=} 1 \}$.
  The derived bound is
  $2|[0,k]| + |[0,na]|$.
  }
\label{fig:xmplincaux}
\end{figure}

\begin{figure}
\vspace{-.3cm}
\begin{lstlisting}[numbers=left]
${\color{blue}%
  \mbox{logical state invariant }
  \{ lg > \log_2(h - l) \} }$
bsearch(x,l,h@,lg@) {
  if (h-l > 1) {
    @assert(lg > 0);@
    m = l + (h-l)/2;
    @lg--;@ if (a[m]>x) h=m; else l=m;
    tick($M_\code{bsearch}$);
    l = bsearch(x,l,h@,lg@);
    tick($-M_\code{bsearch}$);
  } else return l;
}
\end{lstlisting}
\vspace{-.2cm}
\caption{Assisted bound derivation using logical state.
  % We model the stack consumption of the program using tick statements.
  % Before a function call to \code{bsearch}, a tick statement pays for
  % the space used by stack frame of the function.  The space is recoverd
  % after the call.
  We write $\log_2(x)$ for the integer part of
  logarithm of $x$ in base 2.
  The semi-automatically derived bound is $|[0,lg]|$.
  }
\label{fig:xmplbsaux}
\end{figure}

\ifdefined\fullversion
\begin{figure}
\begin{lstlisting}
${\color{blue}
  \mbox{logical state invariant }
  \{ nm = A(0) \} }$
i=1; m=a[0];
while (i < N) {
  if (a[i] > m) {
    @assert(nm > 0); nm--;@
    m = a[i], tick(1);
  } i++, tick(1);
}
\end{lstlisting}
\caption{Example of assisted bound derivation using logical state.
  We write $A(i)$ for
  $\#\{ k \mid i \le k \le N \land \forall\, 0\le j<k.\, a[j] < a[k]\}$.
  The bound derived is $|[0,N]| + |[0,nm]|$.
  }
\label{fig:xmplmax}
\end{figure}
\fi

@@

To compare our tool with existing work, we focused on loop bounds and
use tick statements to counts the number of back edges (i.e., number
of loop iterations) that are followed in the execution of the program
because most other tools only bound this specific cost. In
\pref{fig:compar}, we show the bounds we derived ($\toolname$) together with
the bounds derived by LOOPUS~\cite{SinnZV14} and
Rank~\cite{AliasDFG10}.  We also contacted the authors of SPEED but
have not been able to obtain this tool.
KoAT~\cite{BrockschmidtEFFG14} and PUBS~\cite{AlbertAGPZ12} currently
cannot operate on C code and the examples would need to be manually
translated into a term-rewriting system to be analyzed by these tools.  For
Rank it is not completely clear how the computed bound relates to the C
program since the computed bound is for transitions in an automaton
that is derived from the C code.  For instance, the bound $2+y-x$ that
is derived for \emph{t08} only applies to the first loop in the
program.

@@

\pref{tab:cbench} contains a compilation of the results of our
experiments with the cBench benchmark suite. It shows a representative
list of automatically derived function bounds. In total we analyzed
more than $\totalLoc$ lines of code.  In the LoC column
we not only count the lines of the analyzed function but
also the ones of all the function it calls. \iffull{The line count is
computed on pre-processed C with all unnecessary declarations removed.}{}
We analyzed the functions by giving a cost 1 to all
the back-edges in the control flow (loops, and function calls).
The bounds for the functions \code{ycc\_rgb\_conv} and \code{uv\_decode}
have been inferred with user interaction as described in \pref{sec:anno}.
The most challenging functions for $\toolname$ have unrolled loops
where many variables are assigned.  This stresses our analysis because
the number of LP variables has a quadratic growth in program variables.
Even on these stressful examples, the analysis could finish in less
than 2 seconds.  \iffull{These times can be reduced by adding
manual annotations to indicate $\toolname$ that certain variables do
not have to be tracked.  We could also envision to implement heuristics.}{}
%
For example, the \code{sha\_update} function
is composed of one loop calling two helper functions that in turn
have 6 and 1 inner loops.  In the analysis of the SHA algorithm,
the compositionality of our analysis is essential to get a tight bound
since loops on the same index are sequenced 4 and 2 times without
resetting it.  All other tools derive much larger constant factors.

@@

In this appendix we provide a non-exhaustive catalog of classes of
programs that can be automatically analyzed by our system.  For
simplicity, and to compare our analysis with existing tools, we use
tick statements to count the number of loop iterations and function
calls.  This is the only resource that can be measured by tools like
KoAT~\cite{BrockschmidtEFFG14} and LOOPUS~\cite{SinnZV14}.  Sometimes
we also use (negative) tick statements to discuss features such as
resource restitution.

@@

Example \emph{ex1} from~\cite{GulwaniZ10} specifically focuses on the
code in the \emph{if} statement.  So we
insert \code{tick(1)} inside the if statement to derive a bound on
the number of times the code in the if statement is executed.  Note
that we cannot derive a bound for the whole program since the outer
loop is executed a quadratic number of times.  Nevertheless it is
straightforward to derive a bound on the number of ticks using the
amortized approach: In the if statement we know that $n>0$ and assign
\code{n=n-1}.  So we can use the potential of the interval size $|[0,n]|$
to pay for the tick.

@@

Example \emph{t46} shows a program that uses and returns resources.
Again, we use the function \code{tick} to
describe the resource usage.  The function \code{produce} produces
$|[0,x]|$ resources, that is, in each of the $|[0,x]|$ iterations, it
receives one resource unit.  Similarly, the function \code{consume}
consumes $|[0,y]|$ resources.  The analyzer computes the tight bound
$|[0,y]|$ for the function \code{main}.  This is only possible since
amortized analysis naturally tracks the size changes to the variables
$x$ and $y$, and the interaction between \code{consume} and
\code{produce}.
